### 2011-02-07 Documentation/atropos/ResponseToReviewers.tex

\documentclass[11pt]{article}
\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\y}{{\bf y}}
\renewcommand{\v}{{\bf v}}
\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}

\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\usepackage{verbatim}
\usepackage[usenames]{color}
\usepackage{listings}
\lstset{breaklines=true,basicstyle=\ttfamily,breakindent=0pt}
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\begin{document}
%\maketitle
\noindent
David Kennedy, Ph.D.,\\
Co-Editor-in-Chief, Neuroinformatics\\

\noindent
Dear Dr. Kennedy and Reviewers,\\

\noindent
My coauthors and I would like to thank the editors and reviewers for
handling our manuscript entitled ``An Open Source Multivariate Framework for $n$-Tissue
  Segmentation with Evaluation on Public Data.''  The reviews were extremely helpful in highlighting both
the strengths and weaknesses of the previous version of the paper.  In
summary, the review found the work to be worthy of publication, with
revisions that focused on improving clarity of some key points. We make every effort, in this response letter and the
associated revision, to clarify our work and improve the paper
according to the reviewers' suggestions. We believe that we carefully
revised the manuscript to address all concerns and provide responses
to the reviewers' comments, which you will find below.   We
also elected to add detail regarding the ICM strategy
available in Atropos. Thus, we would
like to resubmit the revised manuscript which we hope will be suitable
for publication. \\

\noindent
The revision of the paper includes:
\begin{enumerate}
\item  A new figure revision highlighting ICM and providing a simple
  starter example for Atropos usage. 
\item Fixes to two typos in the equations. 
\item  Many minor changes in the body of the text.  
\item  The references were altered according to reviewer suggestions.
\end{enumerate}
We express our sincere gratitude to
the editorial staff and to the reviewers for their efforts in
evaluating this paper. \\

\noindent
Sincerely,\\

\noindent
Brian Avants, Ph.D.\\
Penn Image Computing and Science Laboratory \\ 
University of Pennsylvania\\
3600 Market Street, Ste 370\\
Philadelphia, PA 19104
\newpage
\begin{center}
\LARGE{{\bf Response To Reviewers}}
\end{center}
\section{Reviewer 1}
We thank reviewer 1 for his/her careful reading for technical content
and for improving our work.\\
\newline
\newline
{\em 1) Page 5, 2nd line from bottom. What are silver standard
  labellings?}
\newline
\newline
We've changed this line to be more specific in the current revision.
\newline
\newline
Done.  
\newline
\newline
{\em 2) Page 10, Equations 8 - 11. These follow standard
  approaches. However, Equation 11 does not seem consistent with [8]
  and [10]. If delta(x\_i,x\_j)=1 for x\_i=x\_j, then U(x) as defined in
  [10] increases, and p(x) as defined in [8] decreases - hence
  labelling similar regions seems to be penalised. However, things
  seem consistent if equation 11 is changed as in the standard graph
  cuts approaches - e.g. see equation 3 Boykov \& Jolly ICCV 2001.}
\newline
\newline
We appreciate the reviewer catching this error in the equation.  If $x_i = x_j$ then $V=0$ and is
1 otherwise, as is standard in Besag's approach. 
\newline
\newline
{\em 3) Page 10, Equation 11. Computing the value of U(x) only
  considers if neighbours are the same or not. However, as this paper
  deals with segmenting a large number of labels why was an approach
  such as that taken in Noe \& Gee IPMI 2001 not considered -
  i.e. learning the probability of labels being neighbours and using
  these?}
\newline
\newline
This is an interesting point and we intend to investigate this
direction of research in future work.  In fact, some basic
functionality along these lines is available within the guts of the
code but additional comparative research into the myriad partial volume
estimation models that have been proposed is required to determine the
best approach.  Thus, we added comments in the discussion to indicate
our interest in pursuing partial volume models but elect not to pursue
this task for this resubmission. 
\newline
\newline
{\em 4) Page 12, lines 10-14. Authors comment that applying graph cuts
  to more than one label is not globally optimal and that EM is a
  reasonable and efficient alternative. Although this is true,another
  property of graph cuts is that it is fast. How does the EM approach
  compare in terms of speed?}
\newline
\newline
Fast is a relative term and we cannot comment on the speed of a method
for which we do not have an optimal implementation.  
The main computational challenge with graph cuts is memory.  Because one needs to
solve multiple shortest path problems---and store the solution of
each---the potential for memory bloat is high.  Even in EM, memory
concerns emerge when enough labels are used (e.g. in our examples).
In theory, though, EM computations are embarassingly parallel which is
not a property shared by graph-based methods.  As such, one would
expect even an optimally implemented multi-label graph cut method to
lag in speed behind an optimally implemented EM segmentation. 
\newline
\newline
{\em 5) Page 13, Equation 17. This looks like a multivariate equation to me. I'm also not familiar with the form of Equation 19 - why is it different from 17?'}
\newline
\newline
The difference between equations 17 and 19 is that the input to the
data term is a vector (rather than scalar) and the scalar variance
parameter is replaced by the covariance matrix.  This is a standard
extension from univariate to multivariate probability. 
\newline
\newline
{\em 6) Page 17. Penultimate line. "Three likelihood classes have been
  developed, one parametric and one ...". Should this be "Three
  likelihood classes have been developed, one parametric and [two]
  ..."?'}
\newline
\newline
Yes, thank you.  The fix is made.  
\newline
\newline
{\em 7) Page 26 Figure 5. In the sign column the signs for the Left
  and Right Caudate Nucleus and Thalamus should be -ve as according to
  the data in the table, Atropos performed significantly worse than
  the majority vote.}
\newline
\newline
Fixed.  
\newline
\newline
{\em 8) I could not verify the validity of all the references as they were not numbered neither were they in alphabetical order. }
\newline
\newline
{\bf FIXME NICK Standard neuroinformatics style?}
\newline
\newline
The reviewer also mentioned not being able to locate the 19 Hammers
datasets.  It is unfortunate that the brain-development website is not
well annotated.  We suspect the reviewer looked at the IXI data which
is not the Hammers data.  We suggest the reviewer look at the adult
atlas section of the website which, currently, contains the data we
used.  This is noted in the revised text. 
\newline
\newline
We also followed each of the reviewer's suggestions on the structure
and grammar of the text. 
\newline
\newline
We appreciate reviewer 1's time and effort.  
\newline
\newline
\section{Reviewer 2}
We thank reviewer 2 for his/her attention to the manuscript and overall 
positive comments.  We made the reviewer's suggested minor changes in this revision.  
The following comments were addressed: 
\newline
\newline
{\em I was very pleased with the clarity of most of the article,
 particularly with the careful introductory explanations,
 demonstrations of the command-line arguments for Atropos, and the
 excellent review of the literature. I was also happy to see the
 multivariate results in Figure 2, and would have liked to have seen
 accompanying quantitative results. If such results were to assert the
 significant improvement possible when including a complementary
 modality (i.e., without bias), that would be a very nice contribution
 in itself.}
\newline
\newline
We agree with the reviewer that establishing the benefit of
multivariate segmentation in the presence of bias would be
significant.  However, because we only have one multivariate dataset
on which to test this hypothesis, it is difficult to do much more than
provide suggestive results, as we've done here.
\newline
\newline
{\em I have only two concerns with the article. First, there is almost
  no explanation of any of the figures in the body of the text, which
  raises questions about the significance of the findings, such as
  those of Figure 4 ("The results show that the PriorProbabilityMaps
  with w = 0.5 (far right) gives the best performance for all
  tissues."). 
\newline
\newline
The revision adds explanatory material in the body of the text
associated with each figure.  {\bf FIXME}
\newline
\newline
{\em Second, why is it that "In this study, we labeled the brain web (sic) template very sparsely, quickly and crudely and did not expect highly accurate results."
The results depending on these manual prior labels aren't informative
-- they don't give good results, and don't tell us whether the results
would be better if better manual labels were used.}
\newline
\newline
We agree with the reviewer and removed this comment and the evaluation
results.  We added: {\color{red}{We provide this capability to allow the user to implement
    an interactive editing and segmentation loop.  The user may
    run Atropos with sparse manual label guidance, evaluate the
    results, update the manual labels and repeat until achieving the desired
    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.}
\newline
\newline
{\em On page 22: "The advancements introduced with N4 permit such an
  adaptive integration with Atropos." What relevant advances are
  introduced with respect to N3?}
\newline
\newline
{\bf FIXME NICK}
\newline
\newline
{\em Equations: Please remove punctuation following equations, because it came sometimes be distracting. For example, on pg.9, eq.1's comma might also appear to be a prime/apostrophe for the variable y.}
\newline
\newline
Done except in cases where the equation ends a sentence.   
\newline
\newline
Again, we thank each of the three reviewers for their efforts.  We
truly feel that their reading of the paper was excellent and greatly
improved the work.

% \bibliographystyle{elsarticle-harv.bst}
% \bibliography{ftdcbdlong,avantsAll}

\end{document}

### 2011-02-07 Documentation/atropos/atropos.tex
ENH: Getting ready for resubmission to Neuroinformatics
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@763 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
open source segmentation algorithm distributed
with ANTs.%
\footnote{
http://www.picsl.upenn.edu/ANTs
}  The Bayesian formulation of the segmentation problem is 
solved using the Expectation Maximization (EM) algorithm
with the modeling of the class intensities 
 based on either parametric or non-parametric finite mixtures.  
Atropos is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 69 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
problem derived from the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that exercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid (and subdivisions) of the human brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended non-parametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

Several initialization strategies have been proposed to overcome the characteristic
susceptibility of EM algorithms to local optima.   Common low-level initialization steps 
include uniform probability assignment \citep{Wells1996},
Otsu thresholding \citep{Zhang2001}, and K-means clustering
\citep{Pappas1992}.  More sophisticated low-level initialization
schemes include that of \cite{Greenspan2006} in which a dense spatial
distribution of Gaussians is used to capture the complex
neuroanatomical layout with subsequent processing used to conjoin
subsets of such Gaussians belonging to the same tissue classes.  Recently,
anatomically specific strategies for incorporating prior knowledge
into the segmentation solution have included the use of spatial prior
probability maps of the various structures of interest
\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps can also be used to provide an initial segmentation.
Related technological developments model partial volume effects for
increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

@@

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Various techniques
have been proposed to reduce the manual effort required to densely
label a high-resolution neuroimage; one example is the popular
software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces \citep{Dale1999,Fischl1999}.
Advantages of surface representations include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, including Klein's
Mindboggle \citep{Klein2005}, would usually follow an initial
segmentation by a volumetric method such as Atropos.

@@

Our open source segmentation tool, which we have dubbed {\em
Atropos},%
\footnote{Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements a $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid even
in cases of multivariate image data---relevant when more than one 
view of anatomy aids segmentation, as in neonatal brain tissue classification
\citep[e.g.][]{Prastawa2005,Weisenfeld2009}.  
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes with minimal
random access memory requirements, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
also borrow an idea from \cite{Boykov2004} and use sparse spatial
priors to provide initialization and boundary conditions for Atropos
EM segmentation in a semi-interactive manner.  In short, Atropos seeks
to provide a segmentation toolbox that may be modified,
tuned and refined for different use scenarios.

Coupled with the registration \citep{Avants2011} and template building
\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
versatile and powerful software tool which touches multiple aspects of
our brain processing pipeline.  We use Atropos to address brain
extraction \citep{Avants2010a}, grey matter/white matter/cerebrospinal
fluid segmentation, label fusion/propagation and cortical
parcellation.  We also allow Atropos to interact with the recently
developed N4 bias correction software \citep{Tustison2010} in an
adaptive manner.  To further highlight the value of this open source
contribution, we performed a search of software attributes on NITRC
and found that as of November 2010 no stand-alone EM methods are
currently listed.  We also evaluate Atropos performance on two brain
MRI segmentation objectives.  First, three tissue classification.
Second, we test our ability to parcellate the brain into 69
neuroanatomical regions to illustrate the practical value of the
low-memory implementation within this paper.  Although Atropos may be
applied to multivariate data from arbitrary modalities, we limit our
evaluation to tissue-classification in T1 neuroimaging in part due to
the abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

Organization of this work is as follows: we first describe the theory
behind the various components of Atropos while
acknowledging that more theoretical discussion is available elsewhere.  
This is followed by a thorough
discussion of implementation which, though often overlooked, is of
immense practical utility.  We then report evaluation results on
public T1 neuroimaging datasets that have reference gold/silver
standard labelings.  Finally, we provide a discussion of our results
and our open source contribution in the context of the remainder of
this paper and of previous and future work.

@@

Atropos encodes a family of Bayesian segmentation techniques that may
be configured in an application-specific manner.  The theory
underlying Atropos dates back 20$+$ years and is representative of
some of the most innovative work in the field.  Although we synthesize
some of the theoretical work in this section, we recommend to the
interested reader consultation with the literature cited previously
for additional perspective.

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation employ a user-selected observation model defining the
likelihood term and one or more prior probability terms.  The product
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based and/or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation algorithms.

@@

 Atropos optimizes a class
of user-selectable objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{p(\mathbf{y})}, 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail the 
Atropos EM along with choices for the likelihood and prior terms.

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
quantity, we replace the Euclidean distance by by Mahalanobis distance
and define multivariate Gaussian parameters via a mean vector,
$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma}_k$.

>>>>

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
open source segmentation algorithm distributed
with ANTs.%
\footnote{
http://www.picsl.upenn.edu/ANTs
}  The Bayesian formulation of the segmentation problem is 
solved using the Expectation Maximization (EM) algorithm
with the modeling of the class intensities 
 based on either parametric or non-parametric finite mixtures.  
Atropos is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 69 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three-tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 69-class EM segmentation
problem derived from the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that exercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid (and subdivisions) of the human brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  The Expectation-Maximization (EM) framework is
natural \citep{Dempster1977} given the ``missing data'' aspect of this
problem.  The work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended non-parametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

Several initialization strategies have been proposed to overcome the characteristic
susceptibility of EM algorithms to local optima.   Common low-level initialization steps 
include uniform probability assignment \citep{Wells1996},
Otsu thresholding \citep{Zhang2001}, and K-means clustering
\citep{Pappas1992}.  More sophisticated low-level initialization
schemes include that of \cite{Greenspan2006} in which a dense spatial
distribution of Gaussians is used to capture the complex
neuroanatomical layout with subsequent processing used to conjoin
subsets of such Gaussians belonging to the same tissue classes.
Recently, reseachers have begun to rely on spatial prior probability
maps of anatomical structures of interest to encode domain knowledge\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps may also provide an initial segmentation.
Related technological developments model partial volume effects for
increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

@@

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} which aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Various techniques
have been proposed to reduce the manual effort required to densely
label a high-resolution neuroimage; one example is the popular
software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces \citep{Dale1999,Fischl1999}.
Advantages of surface representations include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, including Klein's
Mindboggle \citep{Klein2005}, would usually follow an initial
segmentation by a volumetric method such as Atropos.

@@

Our open source segmentation tool, which we have dubbed {\em
Atropos},%
\footnote{Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements an $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid even
in cases of multivariate image data---relevant when more than one 
view of anatomy aids segmentation, as in neonatal brain tissue classification
\citep[e.g.][]{Prastawa2005,Weisenfeld2009}.  
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes with minimal
random access memory requirements, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
also borrow an idea from \cite{Boykov2004} and use sparse spatial
priors to provide initialization and boundary conditions for Atropos
EM segmentation in a semi-interactive manner.  In short, Atropos seeks
to provide a segmentation toolbox that may be modified,
tuned and refined for different use scenarios.

Coupled with the registration \citep{Avants2011} and template building
\citep{Avants2010} already included in the ANTs, Atropos is a
versatile and powerful software tool which touches multiple aspects of
our brain processing pipeline.  We use Atropos to address brain
extraction \citep{Avants2010a}, grey matter/white matter/cerebrospinal
fluid segmentation, label fusion/propagation and cortical
parcellation.  We also allow Atropos to interact with the recently
developed N4 bias correction software \citep{Tustison2010} in an
adaptive manner.  To further highlight the value of this open source
contribution, we performed a search of software attributes on NITRC
and found that as of November 2010 no stand-alone EM methods are
currently listed.  We also evaluate Atropos performance on two brain
MRI segmentation objectives.  First, three-tissue classification.
Second, we test our ability to parcellate the brain into 69
neuroanatomical regions to illustrate the practical value of the
low-memory implementation within this paper.  Although Atropos may be
applied to multivariate data from arbitrary modalities, we limit our
evaluation to tissue classification in T1 neuroimaging in part due to
the abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

Organization of this work is as follows: we first describe the theory
behind the various components of Atropos while acknowledging that more
theoretical discussion is available elsewhere.  This is followed by a
thorough discussion of implementation which, though often overlooked,
is of immense practical utility.  We then report results on the
BrainWeb and Hammers dataset.  Finally, we
provide a discussion of our results and our open source contribution
in the context of the remainder of this paper and of previous and
future work.

@@

Atropos encodes a family of Bayesian segmentation techniques that may
be configured in an application-specific manner.  The theory
underlying Atropos dates back 20$+$ years and is representative of
some of the most innovative work in the field.  Although we summarize
some of the theoretical work in this section, we recommend that 
the interested reader consult the deep literature in this field for
additional perspective and proofs behind the major concepts.

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation employ a user selected observation model defining the
likelihood term and one or more prior probability terms.  The product
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based and/or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation algorithms.

@@

 Atropos optimizes a class
of user selectable objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{p(\mathbf{y})} 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail the 
Atropos EM along with choices for the likelihood and prior terms.

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
quantity, we replace the Euclidean distance by Mahalanobis distance
and define multivariate Gaussian parameters via a mean vector,
$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma}_k$.

A common technique for the non-parametric variant is to define $p_k$ using Parzen windowing of the sample observation histogram of $\mathbf{y}$, i.e.
\begin{align} \label{eq:nonparam}
  p_k\left(y_i|l_k\right) &= \frac{1}{N_B} \sum_{j=1}^{N_B} \frac{1}{\sqrt{2\pi \sigma_j^2}}\exp\left( \frac{ -(y_i - c_j)^2 }{2\sigma_j^2} \right)
\end{align}
where $N_B$ is the number of bins used to define the histogram of the
sample observations (in Atropos the default is $N_B = 32$) and $c_j$
is the center of the $j^{th}$ bin in the histogram.  $\sigma_j$ is the width of each of the $N_B$ Gaussian kernels.  For multi-modal data in which the number of components of $y_i$ is greater than one, a Parzen window function is constructed for each component.  The likelihood value is determined by the joint probability given by their product.
